---
title: "stage_1_candidate_sp"
author: "Shahar Chaikin"
date: "2025-11-14"
output: html_document
---
Library
```{r}
library(tidyverse)
```

Data entries:
ID_ALL_RAW_DATA Unique BioTIME identifier for record

ABUNDANCE Double representing the abundance for the record (see metadata for details of ABUNDANCE_TYPE

BIOMASS Double representing the biomass for the record(see metadata for details of BIOMASS_TYPE ID_SPECIES Unique identifier linking to the species table

SAMPLE_DESC Concatenation of variables comprising unique sampling event

LATITUDE Latitude of record

LONGITUDE Longitude of record

DEPTH Depthor elevation of record if available

DAY Numerical day of record

MONTH Numerical value of month for record, i.e. January=1

YEAR Year of record

STUDY_ID BioTIME study unique identifier

newID Validated species identifier key

valid_name Highest taxonomic resolution of individual, preferred is genus and species resolution Level of resolution, i.e. ’species’ represented by genus and species taxon Higher level taxonomic grouping, i.e. Fish

#Find candidates
```{r}
biotime_med_fish=read_rds("C:\\Users\\User\\Desktop\\research\\data\\BioTIME 2\\biotime_v2_query_15April25.rds")

m_data=read.csv("C:\\Users\\User\\Desktop\\research\\data\\BioTIME 2\\biotime_v2_metadata_15April25.csv") %>% 
  filter(STUDY_ID%in%biotime_med_fish$STUDY_ID) %>% 
  filter(REALM%in%"Marine")

#Keep marine and species level
biotime_med_fish_marine=biotime_med_fish %>% 
  filter(STUDY_ID%in%m_data$STUDY_ID,
         resolution%in%"species") %>% 
  drop_na(ABUNDANCE)  %>% 
  mutate(h3_id=h3jsr::point_to_cell(
    input = sf::st_as_sf(.,
                         coords = c("LONGITUDE", "LATITUDE"),
                         crs = 4326),
    res = 6))
```

Estimate candidates
```{r}
biotime_cand=biotime_med_fish_marine %>% 
  group_by(h3_id,valid_name) %>% 
  filter(!max(YEAR)==1992) %>% 
  summarise(n=n_distinct(SAMPLE_DESC),
            min_yr=min(YEAR),
            max_yr=max(YEAR),
            year_range=max_yr-min_yr,
            lat_ext=max(LATITUDE)-min(LATITUDE),
            lon_ext=max(LONGITUDE)-min(LONGITUDE)) %>% 
  mutate(pop_id=paste(h3_id,valid_name))

#Species level summary
biotime_cand_sp_level=biotime_cand %>% 
  group_by(valid_name) %>% 
  summarise(n_population=n_distinct(pop_id))
```

#Std. abundances
Grid samples by h3_id
```{r}
#Consider resampling to maintain equal sampling effort
##grid by hexagons
biotime_med_fish_marine_h3_l_ocel=biotime_med_fish_marine %>% 
  mutate(h3_id = h3jsr::point_to_cell(
    input = sf::st_as_sf(.,
                         coords = c("LONGITUDE", "LATITUDE"),
                         crs = 4326),
    res = 5),
    assemblageID=paste(STUDY_ID,h3_id,sep = "_"),
    unique_pop_id=paste(valid_name,STUDY_ID,h3_id,sep = "_")) %>% 
  select(study_id=STUDY_ID,
         sample_id=SAMPLE_DESC,
         lat=LATITUDE,
         long=LONGITUDE,
         h3_id,
         assemblageID,
         unique_pop_id,
         year=YEAR,
         valid_name,
         abund=ABUNDANCE) %>% 
  filter(valid_name%in%"Leucoraja ocellata")


#How many sp and samples per study?
check_samples <- biotime_med_fish_marine_h3_l_ocel %>% 
  group_by(study_id,h3_id,assemblageID, year) %>% 
  summarise(n_samples = n_distinct(sample_id), n_species = n_distinct(valid_name))
check_samples$n_samples %>% hist
check_samples$n_samples %>% quantile()
```

Resample
Consider resampling samples (i.e., SAMPLE_DESC) within assemblageID (study within hexagon) to maintain equale sampling effort over the years. Chose the cutoff of how many samples per year to sample.
```{r}
#remove populations that were annualy poorly sampled (<3)
biotime_med_fish_marine_h3_poorly=biotime_med_fish_marine_h3_l_ocel %>% 
  group_by(unique_pop_id,year) %>% 
  filter(n_distinct(sample_id)>=3) %>% 
  ungroup()

check_samples2 <- biotime_med_fish_marine_h3_poorly %>% 
  group_by(assemblageID,unique_pop_id,study_id,h3_id,year) %>% 
  summarise(n_samples = n_distinct(sample_id), n_species = n_distinct(valid_name))

#resample within assamblage IDs according to these values
check_samples2_min=check_samples2 %>% 
  group_by(unique_pop_id) %>% 
  filter(n_samples==min(n_samples)) %>% 
  distinct(unique_pop_id,n_samples)

#How many assamblages?
biotime_med_fish_marine_h3_poorly$assemblageID %>% n_distinct()
```

#Resample for equal samp_effort:
Defining the Fixed Effort (m): The code first calculates the required number of distinct samples, m, for every unique_pop_id within an assemblageID.

It finds the annual count of distinct sample_ids for that group.

It then takes the minimum of these counts across all years. This minimum value, m, becomes the target effort.

Enforcing the Fixed Effort: In the resampling loop, the do() function, combined with sample(..., size = m, replace = FALSE), is executed for every assemblageID/unique_pop_id/year group:

group_by(assemblageID, unique_pop_id, year) ensures the operation is applied year-by-year for each population.

sample(distinct_samples, size = m, ...) randomly selects exactly m distinct sample_ids from the available samples for that year.

By forcing the selection of m distinct samples for every year, the code eliminates the variation in effort that was caused by differences in sampling intensity between years.

In essence, the entire procedure reduces the higher-sampled years down to the effort level of the lowest-sampled year, thus creating a balanced dataset where the sampling effort (i.e., the number of distinct samples used) is constant across all years for a given population.

```{r}
# Load necessary libraries
library(rlang) # Required for non-standard evaluation

# Assuming the data is already loaded and named 'biotime_med_fish_marine_h3_poorly'

# Set the number of iterations
n_iterations <- 100

# 1. Calculate the minimum annual sample count per unique_pop_id within an assemblageID
min_samples_df <- biotime_med_fish_marine_h3_poorly %>%
  # Group by the required levels for calculating the minimum
  group_by(assemblageID, unique_pop_id, year) %>%
  # Count the number of distinct samples for each group
  summarise(n_samples = n_distinct(sample_id), .groups = "drop") %>%
  # Re-group to find the minimum sample count per unique_pop_id across all years
  group_by(assemblageID, unique_pop_id) %>%
  # Find the minimum sample count for this unique_pop_id across all its years
  # This is the 'fixed sampling effort' (m)
  summarise(min_n_samples = min(n_samples), .groups = "drop")

# 2. Resampling and Iteration

# A list to store the results of each iteration
resampled_data_list <- list()

# A tibble to track the number of times each row (species record) was used
# Initialize with a count of 0 for every row in the original data
n_used_tracker <- biotime_med_fish_marine_h3_poorly %>%
  # Assign a unique row_id for easy tracking
  mutate(row_id = row_number()) %>%
  select(row_id) %>%
  mutate(n_used = 0)

# Function to perform one iteration of resampling
resample_iteration <- function(data, min_counts) {
  # Join the data with the pre-calculated minimum sample counts
  data_with_min <- data %>%
    left_join(min_counts, by = c("assemblageID", "unique_pop_id"))

  # Resample according to 'min_n_samples'
  resampled_df <- data_with_min %>%
    group_by(assemblageID, unique_pop_id, year) %>%
    # Use 'slice_sample' to randomly select 'min_n_samples' distinct sample_ids
    # from the current group (assemblageID, unique_pop_id, year).
    # 'replace = FALSE' ensures sampling without replacement.
    # The 'n' argument is set dynamically by the 'min_n_samples' column.
    # NOTE: The selection is based on distinct 'sample_id', and all rows
    # associated with the selected 'sample_id' are kept *if* the group size
    # for the unique_pop_id, year is > 0.
    # A more rigorous selection based on *distinct* sample_id first:
    do({
        # Get distinct sample_ids for the current group
        distinct_samples <- unique(.$sample_id)
        # Get the required minimum count (m)
        m <- first(.$min_n_samples)

        # Ensure m is not NA and is positive
        if (is.na(m) || m <= 0) {
            # Skip if no valid minimum count exists
            return(tibble())
        }

        # Select 'm' distinct sample_ids without replacement
        selected_sample_ids <- sample(
            distinct_samples,
            size = min(m, length(distinct_samples)), # Use min to avoid error if m > available
            replace = FALSE
        )

        # Filter the original data rows for these selected sample_ids
        # Return all rows corresponding to the selected sample_ids
        filter(., sample_id %in% selected_sample_ids)
    }) %>%
    ungroup() %>%
    # Remove the temporary column
    select(-min_n_samples)

  return(resampled_df)
}

# Run the iterations and track usage
for (i in 1:n_iterations) {
   # SET SEED HERE for reproducibility of the random sampling step
  set.seed(42 + i) # Use a seed that changes but is deterministic based on the iteration

  # 1. Perform resampling
  resampled_iteration <- resample_iteration(biotime_med_fish_marine_h3_poorly %>% mutate(row_id = row_number()), min_samples_df)

  # 2. Store the result (for mean abundance calculation)
  # Add an iteration identifier
  resampled_data_list[[i]] <- resampled_iteration %>%
    mutate(iteration = i)

  # 3. Update the usage tracker ('n_used')
  # Get the row_ids that were selected in this iteration
  selected_row_ids <- resampled_iteration$row_id

  # Update the counts in the tracker
  n_used_tracker <- n_used_tracker %>%
    mutate(
      n_used = ifelse(row_id %in% selected_row_ids, n_used + 1, n_used)
    )
}

# 3. Final Aggregation

# Combine all iteration results
all_resampled_data <- bind_rows(resampled_data_list)

# Calculate the mean abundance (mean of 'abund' column)
mean_abundance_df <- all_resampled_data %>%
  # Group by all identifiers and year
  group_by(h3_id, assemblageID, unique_pop_id, lat, long, year) %>%
  # Calculate the mean abundance across all iterations for this group
  summarise(
    mean_abundance = mean(abund, na.rm = TRUE),
    .groups = "drop"
  )

# Join the mean abundance with the usage count
# Note: The usage count 'n_used' is per original row, which doesn't directly
# map to the aggregated groups. Instead, we can calculate the *total* times a
# unique_pop_id was sampled *across all iterations*.

# Calculate n_used (total selection count across all iterations for each unique_pop_id)
# This is a bit ambiguous in the prompt: "n_used (which is the number of times
# a certain species was selected across all iterations)". Assuming it means the
# *total number of observations (rows) used* for that final aggregated group.
n_used_aggregated <- all_resampled_data %>%
  # Group by the final aggregation levels
  group_by(h3_id, assemblageID, unique_pop_id, lat, long, year) %>%
  # Count the total number of rows used across all 100 iterations
  summarise(
    n_used = n(),
    .groups = "drop"
  )

# Final Dataset Creation
final_dataset <- mean_abundance_df %>%
  # Join with the aggregated n_used count
  left_join(n_used_aggregated, by = c("h3_id", "assemblageID", "unique_pop_id", "lat", "long", "year"))

#Population summary
pop_summary=final_dataset %>% 
  group_by(unique_pop_id) %>% 
  summarise(years_sampled=n_distinct(year),
            min_yr=min(year),
            max_yr=max(year),
            ann_range=(max(year)-min(year))+1,
            coverage=years_sampled/ann_range) %>% 
  filter(years_sampled>=3)
#clean based on year cutoff
final_dataset_cutoff=final_dataset %>% 
  filter(unique_pop_id%in%pop_summary$unique_pop_id)

final_dataset_cutoff$unique_pop_id %>% n_distinct()
```

Population summary
```{r}
final_dataset_cutoff %>% summary
```

#Estimate slope
```{r}
library(broom)
# Fit linear models per unique_pop_id
slopes_df <- final_dataset_cutoff %>%
  group_by(unique_pop_id,h3_id) %>%
  nest() %>%
  mutate(
    model = map(data, ~ lm(mean_abundance ~ year, data = .x)),
    tidy_out = map(model, ~ tidy(.x, conf.int = TRUE))) %>%
  unnest(tidy_out) %>%
  filter(term == "year") %>%
  select(unique_pop_id, estimate, std.error, conf.low, conf.high) %>%
  mutate(weights=(1/std.error)) %>% 
  rename(
    slope = estimate,
    slope_se = std.error,
    slope_ci_low = conf.low,
    slope_ci_high = conf.high) %>% 
  left_join(pop_summary %>% 
              select(unique_pop_id,min_yr,max_yr),
            by = "unique_pop_id") %>% 
  ungroup() %>% 
  mutate(centroids=h3jsr::cell_to_point(h3_id),
    lon = sf::st_coordinates(centroids)[, 1],
    lat = sf::st_coordinates(centroids)[, 2]) %>% 
  select(-centroids)
```

Summary
```{r}
ggplot()+
  geom_density(data=slopes_df,
                 aes(x=slope),
               fill="grey")+
  geom_vline(xintercept = slopes_df %>% 
               summarise(mean_slope=mean(slope)) %>% 
               pull())+
  geom_vline(xintercept = 0,
             linetype="dashed")+
  theme_bw()
```

Slopes by latitude
```{r}
lm_alope_lat=lm(data=slopes_df,
                formula = slope~lat,
                weights=weights)
lm_alope_lat %>% summary


ggplot(slopes_df)+
  geom_point(aes(x=lat,y=slope,size=1/slope_se))+
  theme_classic()+
  labs(y="Slope",x="Latitude",size="Inverse SE")+
  geom_smooth(formula = y~x,
              aes(x=lat,y=slope),
              method = "lm")
```

Map slopes
```{r}
slopes_df_spat=slopes_df %>% 
  mutate(poly=h3jsr::cell_to_polygon(h3_id))
```

map
```{r}
# Load world map
world <- rnaturalearth::ne_countries(scale = "large", returnclass = "sf")

# Calculate bounding box with buffer
bbox <- slopes_df_spat %>%
  summarise(
    xmin = min(lon) - 2,
    xmax = max(lon) + 2,
    ymin = min(lat) - 2,
    ymax = max(lat) + 2
  )

# Extract bounding box values
xlim_vals <- c(bbox$xmin[[1]], bbox$xmax[[1]])
ylim_vals <- c(bbox$ymin[[1]], bbox$ymax[[1]])

# Plot
# Convert to sf and set geometry column
slopes_df_spat_sf <- st_as_sf(slopes_df_spat, sf_column_name = "poly")

slope_map_p=ggplot() +
  geom_sf(
    data = slopes_df_spat_sf,
    aes(fill = slope)
  ) +
    geom_sf(data = world, fill = "gray90", color = "white") +
    coord_sf(
    xlim = xlim_vals,
    ylim = ylim_vals,
    expand = FALSE
  ) +
  theme_bw() +
  labs(
    x = "Longitude",
    y = "Latitude",
    fill = "Slope",
    title = "GBIF Occurrences over Time"
  )+
  scale_fill_gradient2(high="blue",
                       mid="white",
                       low="red",midpoint = 0)


ggsave(plot = slope_map_p,filename = "slope_map_p.png",device = "jpeg",width = 20,height = 15,units = "cm")
```

