---
title: "stage_7_thermal_position"
author: "Shahar Chaikin"
date: "2025-11-15"
output: html_document
---
In this code we will estimate the thermal position of each population used
Input:
1) GBIF occurrence (cleaned)
2) Temperature raster (climate)
3) Suitability overtime (based on SDM)
4) Abundance change over time (based on abundance time series lm)

library
```{r}
library(tidyverse)    # essential for data manipulation (dplyr, ggplot2, tidyr, readr)
library(raster)       # for working with raster data (e.g., netcdf temperature files)
library(h3jsr)        # r bindings for h3 geospatial indexing system
library(sf)           # simple features for spatial data handling
library(DataExplorer) # for data exploration and missingness plots (optional)
```

#1) Load spatial thinned occurrence data
```{r}
sp_thin_data=bind_rows(read.csv("raja_gbif_lite_thin_hexs_5thinned.csv")) %>% 
  ungroup()

sp_thin_data_coords=sp_thin_data %>% 
  ungroup() %>% 
  dplyr::select(lat,
         lon) %>% 
  mutate(h3_id=h3jsr::point_to_cell(
    input = sf::st_as_sf(.,
                         coords = c("lon", "lat"),  # Fixed column names
                         crs = 4326), res = 5))
```

#2) Load seabed temperature raster
this section loads the seabed temperature raster data for estimating species' thermal ranges. The data is sourced from bio-oracle.
```{r}
# Load the netCDF file as a raster object
seabed_temp_at_mean_depth <- "env_rast_croped_marine_ecore/btemp_rast.tif" 
temp_raster <-read_rds("res_5_env/bottomT_df_h3_res_5_2.rds") %>% 
  group_by(h3_id) %>% 
  summarise(mean_bottomt=mean(btemp_mean),
            median_bottomt=median(btemp_mean)) %>% 
  mutate(poly=h3jsr::cell_to_polygon(h3_id))
```

#3) Overlap occurrence records with seabed temperatures.
this step extracts the seabed temperature values corresponding to each species occurrence record. This is a step to link populations to their experienced thermal conditions. Records with missing temperature values (e.g., points on land or outside the raster's extent) are removed.
```{r}
# Extract temperature values from the raster at the specified coordinates
extracted_seabedtemp <- sp_thin_data_coords %>% 
  left_join(temp_raster %>% 
              dplyr::select(1,2,3),
            by="h3_id")

sp_thin_data=extracted_seabedtemp %>% 
  drop_na()
```

#4) Create per species seabed temperature summary
this section calculates key summary statistics of seabed temperature for each species based on their occurrence records.
```{r}
sp_temp_data=sp_thin_data %>% 
  drop_na(mean_bottomt) %>% 
  ungroup() %>% 
  summarise(n_occ=n(),
            min_seabed_t=min(mean_bottomt),
            mean_seabed_t=mean(mean_bottomt),
            median_seabed_t=median(mean_bottomt),
            max_seabed_t=max(mean_bottomt),
            mid_point=(max_seabed_t+min_seabed_t)/2,
            thermal_range=max_seabed_t-min_seabed_t) %>% 
  mutate(across(where(is.numeric), round, digits = 3))
```

#5) Estimate the center temperature indices within each hexagon
this section estimates the mean seabed temperature within each h3 hexagonal cell that defines a "population". this environmental context for the populations will be used to compare against the species' overall thermal range
```{r}
# Calculate summary statistics
polygon_values_summary <- temp_raster %>%
  group_by(h3_id) %>% 
  summarize(
    mean_temp = mean(mean_bottomt , na.rm = TRUE) %>% 
      round(digits = 3), 
    median_temp = median(mean_bottomt , na.rm = TRUE) %>% 
      round(digits = 3))

#Add info
polygon_values_summary=polygon_values_summary %>% 
  ungroup() %>% 
  dplyr::select(h3_id,h3_mean_temp=mean_temp,h3_median_temp=median_temp) %>% 
  mutate(h3_centroid=h3jsr::cell_to_point(h3_id),
         h3_center_lon = as.numeric(str_extract(h3_centroid, "-?\\d+\\.\\d+")),
         h3_center_lat = as.numeric(str_extract(h3_centroid, "\\s+(-?\\d+\\.\\d+)"))) %>%
  dplyr::select(-h3_centroid) %>% 
  drop_na()
#write.csv(polygon_values_summary,"h3_id_seabed_temp.csv",row.names = F)
```

#6) Calculate thermal position index
this section combines the species-level thermal information with the h3-level data to calculate a "seabed thermal position" index for each population. this index quantifies a population's thermal position relative to its species' overall thermal range.
```{r}
#Populations info
pop_info=polygon_values_summary
#Calculate thermal position - seabed
thermal_pos= pop_info  %>%
  mutate(species="Leucoraja ocellata") %>% 
  left_join(sp_temp_data %>%
  mutate(species="Leucoraja ocellata"),
            by="species") %>% 
  group_by(h3_id) %>% 
  mutate(
    seabed_thermal_pos=
      ((h3_median_temp-median_seabed_t)/(max_seabed_t-mid_point)) %>% 
           round(digits=3)) %>% 
  drop_na(n_occ)

thermal_pos$species %>% n_distinct()
thermal_pos$seabed_thermal_pos %>% quantile(na.rm = T)
#write.csv(thermal_pos,"thermal_pos.csv",row.names = F)
DataExplorer::plot_missing(thermal_pos)
```
